---
title: "vignette"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{vignette}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(Climatastic)
data("station_data")
data("daily_weather_data")
```

*1. Make a map of the average temperature at each station for the month
of March 2024.*

```{r}
stations <- station_data$station_name

mar24_temp_avgs <- rep(NA, length(stations))
i = 1
start = as.Date("2024-03-01")
end = as.Date("2024-03-31")

for(station_name in stations){
  station_weather <- 
    daily_weather_data[daily_weather_data["station_name"] == station_name, ]
  time_series <- extract_time_series(station_weather, start, end)
  mar24_temp_avgs[i] <- mean(time_series$T_DAILY_AVG, na.rm = TRUE)
  i <- i + 1
}

#some stations do not have data in 2024 or have NA's for all of March '24
df_mar24 <- na.omit(cbind(station_data, mar24_temp_avgs))
names(df_mar24)[6] <- "temp_avg"

df_mar24$LONGITUDE <- as.numeric(df_mar24$LONGITUDE)
df_mar24$LATITUDE <- as.numeric(df_mar24$LATITUDE)

fields::bubblePlot(df_mar24$LONGITUDE, df_mar24$LATITUDE, df_mar24$temp_avg, col = fields::tim.colors(64), xlab = "Longitude", ylab = "Latitude", legend.lab = "Avg Temperature (Celsius)", main = "Average Temperature at Each Station for March 2024")
```

2.  *Fit a spatial model and plot an interpolated map of average
    temperatures for March 2024. Consider including elevation in your
    model.*

```{r}
grid_pts <- create_grid_pts()

library(elevatr) #get elevation data

#first for station data for march 24
mar24_locations <- df_mar24[,c("LONGITUDE", "LATITUDE")]
names(mar24_locations) <- c("x", "y")
mar24_elevations <- get_elev_point(mar24_locations, prj = 4326, src = "aws")

station_covariates <- cbind(mar24_locations, mar24_elevations$elevation)

#now for all grid_pts - Problem: grid points have 335 NA's
grid_locations <- grid_pts[,c("LONGITUDE", "LATITUDE")]
names(grid_locations) <- c("x", "y")
grid_elevations <- get_elev_point(grid_locations, prj = 4326, src = "aws")

grid_covariates <- cbind(grid_locations, grid_elevations$elevation)

```

```{r}
interpolations <- interpolate_grid_pts(
  station_locs = df_mar24[,c("LONGITUDE", "LATITUDE")],
  station_covariates = station_covariates,
  interpolation_data = df_mar24[["temp_avg"]], 
  grid_pts = grid_pts[,c(1,2)],
  grid_covariates = grid_covariates)

plot_interpolations(interpolations, grid_pts, legend.lab = "Average Temperatures (Celsius)", main = "Interpolated Average Temperatures for March 2024")
```

3.  *Estimate the warmest and coldest day of the year for each station,
    and plot those days on two maps. Think carefully about how to
    represent the days numerically.*

*In your report, describe the statistical analysis that you used for
estimating the warmest and coldest days at each station, including
writing down any statistical models in mathematical notation. Be sure to
define all your symbols and assumptions.*

*Interpolate maps of the warmest and coldest days, and plot the
interpolated maps of warmest and coldest days.*

```{r}
stations <- station_data$station_name
lowest_highest <- matrix(NA, length(stations), 2)
colnames(lowest_highest) <- c("lowest", "highest")

for(i in 1:length(stations)){
  station <- stations[i]
  station_daily_data <- daily_weather_data[daily_weather_data$station_name == station, ]
  cycle <- yearly_cycle(station_daily_data) #this gives us tibble of yearly cycle
  lowest_highest[i, "lowest"] <- which.min(cycle$expected_avg_temp)
  lowest_highest[i, "highest"] <- which.max(cycle$expected_avg_temp)
 
}
lowest_highest

station_data
```

Statistical Analysis:

To estimate the warmest and coldest days at each station, we first
create a model to predict the cyclical temperature based on all the
provided data for that specific station. Our model is as follows:
$y_i = sin (2 *\pi *x_i / 365.25) + cos(2* \pi *x_i / 365.25) + \epsilon_i$

Here, $x_i$ represents the day of the year since January 1st, $y_i$
represents the average daily temperature (T_DAILY_AVG). We use the trig
functions to account for temperature fluctuations in a cyclical manner
throughout the year, using a period of $365.25$ to represent the number
of days in year, while also accounting for leap years.

Note that before we trained the model, we removed all instances of leap
years to ensure that all March 1 (and all days afterwards) data points
are considered together in their expected average temperature.

Afterwards, we use our fitted model to predict the expected daily
average temperature for each day (1 through 365) for that specific
station. We did this by obtaining the sine and cosine terms for each day
(1 through 365).

To then obtain the warmest and coldest day in the cycle, we simply
extracted the minimum and maximum of each cycle.

```{r}
grid_pts <- create_grid_pts()

#interpolate lowest
interpolations_lowest <- interpolate_grid_pts(station_locs = station_data[,c("LONGITUDE", "LATITUDE")], 
                     interpolation_data = lowest_highest[, "lowest"], 
                     grid_pts = grid_pts[,c(1,2)])

plot_interpolations(interpolations_lowest, grid_pts, legend.lab = "Day of Year Since Jan. 1", main = "Coldest Days at Each Station")
```

```{r}
#interpolate highest
interpolations_highest <- interpolate_grid_pts(station_locs = station_data[,c("LONGITUDE", "LATITUDE")], 
                     interpolation_data = lowest_highest[, "highest"], 
                     grid_pts = grid_pts[,c(1,2)])
plot_interpolations(interpolations_highest, grid_pts, legend.lab = "Day of Year Since Jan. 1", main = "Warmest Days at Each Station")

```

4.  *Make a single plot of the estimated yearly cycles for 10 different
    stations, highlighting a diversity of climates around the contiguous
    USA. Your plot should clearly indicate which cycle is from which
    station.*

```{r}
ten_stations <- c(53878, 64758, 53152, 23907, 63856, 4136, 54808, 94996, 94075, 53960)
#for each station id: get the name and state
#NC_Asheville_13_S, NY_Ithaca_13_E, CA_Santa_Barbara_11_W, TX_Austin_33_NW, GA_Brunswick_23_S, WA_Spokane_17_SSW, IL_Champaign_9_SW, NE_Lincoln_11_SW, CO_Boulder_14_W, LA_Lafayette_13_SE
ten_station_name = c("NC_Asheville_13_S", "NY_Ithaca_13_E", "CA_Santa_Barbara_11_W", "TX_Austin_33_NW", "GA_Brunswick_23_S", "WA_Spokane_17_SSW", "IL_Champaign_9_SW", "NE_Lincoln_11_SW", "CO_Boulder_14_W", "LA_Lafayette_13_SE")

i <- 1
cycles <- matrix(NA, nrow = 365, ncol = 10)

for (id in ten_stations){
  station_daily_data <- daily_weather_data[daily_weather_data$WBANNO == id, ]
  cycles[,i] <- yearly_cycle(station_daily_data)[["expected_avg_temp"]]
  i = i + 1
}

colors <- rainbow(10)
matplot(1:365, cycles, type = "l", col = colors, xlab = "Days", ylab  = "Expected Average Temperature", main = "Estimated Yearly Cycle for 10 different stations", cex  = 1, .main = 0.9)
legend(x = "top", legend = ten_station_name, col=colors, 
       lwd = 10, cex = 0.5, ncol = 5, xpd = TRUE, inset = c(1, 1.1))

```

5.  *Estimate the trend over the years for each station, in units of
    degrees Celsius per year, and plot the trend values on a map.
    Indicate visually on your map which of the trends are statistically
    significant. In your report, write the statistical model that you
    used in mathematical notation. Be sure to define all your symbols
    and assumptions.*

    *Interpolate the estimated trends to a grid, and plot them on a map.
    For the interpolations, you may consider using only the trend
    estimates whose standard errors are sufficiently small.*

```{r}
stations <- station_data$station_name
trend_estimates <- matrix(NA, nrow = length(stations), ncol = 3)
colnames(trend_estimates) <- c("Coefficient", "Std. Error", "P-value")

i <- 1
for (station in stations){
  station_daily_data <- daily_weather_data[daily_weather_data$station_name == station, ]
  
  trend_estimates[i, ] <- trend_estimate(station_daily_data)
  i = i + 1
}

#For my plot, I will disregard the SA Tiksi station, as that is in Russia and 
# causes the map to zoom out of all other US locations
not_SA <- which(stations != "SA_Tiksi_4_SSE")
trend_estimates_not_SA <- trend_estimates[not_SA,]

#indicating statistical significance
significant_stations <- which(trend_estimates_not_SA[,3] < 0.05)

trend_range <- range(trend_estimates_not_SA[, "Coefficient"]) 
#ensures that both of the following plots have the same scale

#significant station are circles
#non-significant stations are squares
long_numeric <- as.numeric(station_data[not_SA,]$LONGITUDE)
lat_numeric <- as.numeric(station_data[not_SA,]$LATITUDE)


fields::bubblePlot(long_numeric[significant_stations], 
                   lat_numeric[significant_stations], 
                   trend_estimates_not_SA[significant_stations,"Coefficient"], 
                   col = fields::tim.colors(64), 
                   xlab = "Longitude", ylab = "Latitude", 
                   legend.lab = "Trend (Celsius per Year)", 
                   main = "Trend Estimate for Each Station", 
                   zlim = trend_range)

fields::quilt.plot(long_numeric[-significant_stations], 
                  lat_numeric[-significant_stations], 
                  trend_estimates_not_SA[-significant_stations,"Coefficient"], 
                  col = fields::tim.colors(64),
                  zlim = trend_range, add = TRUE, size = 2)
legend("topright", legend = c("Significant", "Not Significant"), 
       pch = c(16, 15), col = "black")

#do another plot without those two in alaska that mess up our scale
```

Statistical Analysis:
Assumptions: we want the linear trend, but in reality the trend is not linear when examining change in temperature since a certain date, as it is best represented by sine and cosine.


```{r}
#Interpolation

#We will consider the standard errors that are sufficiently small to be the ones below 0.1.

is_small_se <- which(trend_estimates[,"Std. Error"] < 0.1)
station_data[is_small_se, c("LONGITUDE", "LATITUDE")]

trend_small_se <- trend_estimates[is_small_se, "Coefficient"]

grid_pts <- create_grid_pts()

interpolations_trend <- interpolate_grid_pts(
  station_locs = station_data[is_small_se, c("LONGITUDE", "LATITUDE")],
  interpolation_data = trend_small_se,
  grid_pts = grid_pts[,c(1,2)]
)

plot_interpolations(interpolations_trend, grid_pts, legend.lab = "Trend (Celsius per Year)", main = "Trend Interpolations")
```

6.  *Find a reputable source for the average temperature trend in the
    contiguous USA over the past 20 years, and compare your results to
    the source’s.*

https://www.epa.gov/climate-indicators/climate-change-indicators-us-and-global-temperature
https://www.climate.gov/news-features/blogs/beyond-data/mapping-us-climate-trends
https://www.weather.gov/media/slc/ClimateBook/Annual%20Average%20Temperature%20By%20Year.pdf
https://ntrs.nasa.gov/api/citations/20180001314/downloads/20180001314.pdf

