---
title: "vignette"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{vignette}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(Climatastic)
```

*1. Make a map of the average temperature at each station for the month
of March 2024.*


```{r}
data("station_data")
data("daily_weather_data")

stations <- station_data$station_name

mar24_temp_avgs <- rep(NA, length(stations))
i = 1
start = as.Date("2024-03-01")
end = as.Date("2024-03-31")

for(station_name in stations){
  station_weather <- 
    daily_weather_data[daily_weather_data["station_name"] == station_name, ]
  time_series <- extract_time_series(station_weather, start, end)
  mar24_temp_avgs[i] <- mean(time_series$T_DAILY_AVG, na.rm = TRUE)
  i <- i + 1
}

#some stations do not have data in 2024 or have NA's for all of March '24
df_mar24 <- na.omit(cbind(station_data, mar24_temp_avgs))
names(df_mar24)[6] <- "temp_avg"
df_mar24

df_mar24$LONGITUDE <- as.numeric(df_mar24$LONGITUDE)
df_mar24$LATITUDE <- as.numeric(df_mar24$LATITUDE)

maps::map("usa")
fields::quilt.plot(df_mar24$LONGITUDE, df_mar24$LATITUDE, df_mar24$temp_avg, add = TRUE, xlab = "Longitude", ylab = "Latitude", legend.lab = "Avg Temperature")
#try to make the squares smaller if we have time
```


2.  *Fit a spatial model and plot an interpolated map of average
    temperatures for March 2024. Consider including elevation in your
    model.*
    
```{r}
grid_pts <- create_grid_pts()

#get elevation data
library(elevatr)

#first for station data for march 24
mar24_locations <- df_mar24[,c("LONGITUDE", "LATITUDE")]
names(mar24_locations) <- c("x", "y")
mar24_elevations <- get_elev_point(mar24_locations, prj = 4326, src = "aws")

station_covariates <- cbind(mar24_locations, mar24_elevations$elevation)

#now for all grid_pts - Problem: grid points have 335 NA's
grid_locations <- grid_pts[,c("LONGITUDE", "LATITUDE")]
names(grid_locations) <- c("x", "y")
grid_elevations <- get_elev_point(grid_locations, prj = 4326, src = "aws")

grid_covariates <- cbind(grid_locations, grid_elevations$elevation)

```

    
```{r}
interpolations <- interpolate_grid_pts(
  station_locs = df_mar24[,c("LONGITUDE", "LATITUDE")],
  station_covariates = station_covariates,
  interpolation_data = df_mar24[["temp_avg"]], 
  grid_pts = grid_pts[,c(1,2)],
  grid_covariates = grid_covariates)

plot_interpolations(interpolations, grid_pts)
```
    
3.  *Estimate the warmest and coldest day of the year for each station,
    and plot those days on two maps. Think carefully about how to
    represent the days numerically. In your report, describe the
    statistical analysis that you used for estimating the warmest and
    coldest days at each station, including writing down any statistical
    models in mathematical notation. Be sure to define all your symbols
    and assumptions. 2 Interpolate maps of the warmest and coldest days,
    and plot the interpolated maps of warmest and coldest days.*
4.  *Make a single plot of the estimated yearly cycles for 10 different
    stations, highlighting a diversity of climates around the contiguous
    USA. Your plot should clearly indicate which cycle is from which
    station.*
5.  *Estimate the trend over the years for each station, in units of
    degrees Fahrenheit per year, and plot the trend values on a map.
    Indicate visually on your map which of the trends are statistically
    significant. In your report, write the statistical model that you
    used in mathematical notation. Be sure to define all your symbols
    and assumptions. Interpolate the estimated trends to a grid, and
    plot them on a map. For the interpolations, you may consider using
    only the trend estimates whose standard errors are sufficiently
    small.*
6.  *Find a reputable source for the average temperature trend in the
    contiguous USA over the past 20 years, and compare your results to
    the source’s.*
